---
title: "Mixed Models 2"

execute: 
  warning: false
  message: false
---

Some studies might need a **generalised mixed model**. Others might have random effects that are **crossed** or **nested**. 

Understanding these concepts will mean you will be aware when they apply to your own studies.

<br>



# Generalised mixed models

Sometimes the response (dependent) variable is not continuous but a categorical variable. Just as you can use alternative **generalized** models instead of lm such as logistic or poisson, you can use other **generalised mixed** models.

For example, imagine each of our dragons was scored as passing or failing the intelligence test making the response `passFail` binary. Then you would use a **binary logistic mixed model**. 

## Dragons again

<br>

**Challenge part 1**

Find the function and the code for a binominal mixed model. Adapt the code to run an analysis on this [binomial dragon data](./data/dragons_binomial.csv) where the response variable `passFail` is if the dragon passed (1) or failed (0) the IQ test. The fixed effect is `bodyLength` and random effect is `mountainRange`. 

```{r}
#| echo: false
#| eval: false
dragons2 <- read.csv(file = "data/dragons_binomial.csv")
library(lme4)
```

<button type="button" class="btn btn-success">

Consider if you need to check for assumptions.  
What numbers from the output are you going to use to interpret the results? You might want to remind yourself of the binomial response section of the linear model lesson. 
Consider comparing full and null models in an anova or comparing AIC values.  

</button>

```{r}
#| echo: false
#| eval: false
model <- glmer(passFail ~ bodyLength + (1|mountainRange), data = dragons2, family = binomial)
# To check for linearity they may try to bootstrap the analysis to get a better graph like in the binomial generalised linear model.
summary(model) # for glm esp binomial p values are unreliable but if look at the estimate for bodylength is pretty much 0.
anova(model) # gives F value but not p. Possibly because can't really get p values for mixed models.
full.model <- glmer(passFail ~ bodyLength + (1|mountainRange), data = dragons2, family = binomial)
reduced.model <- glmer(passFail ~ 1 + (1|mountainRange), data = dragons2, family = binomial)
anova(full.model, reduced.model) # only 2 units difference between AIC and p values is 0.9469 so not effect
```

<br>

**Challenge part 2**

Once you have run the analysis and interpreted the results, write out in your script how you would report this in words. 

<br>

**Challenge part 3**

Decide what a suitable graph would be and create one making sure it's formatted.


<br>

## Lizards eating

Imagine an experiment where lizards in individual tanks are observed everyday for 14 days to see if they eat or not (`lizards$eat`). Some are in a control group with dead prey and some are in a treatment group with live prey. There are males and females. [lizard data](./data/lizard.csv)

```{r}
lizards <- read.csv(file = "data/lizard.csv")
head(lizards)
```


Identify why this needs a mixed model and therefore what the random variable is.

<details>
  <summary>**Answer**</summary>

We need to control for the differences among individual animals because we have repeated measures. In other words the measurements are not independent because some of them come from the same individual lizard. Therefore, `lizard` is the random variable.
</details>

<br>

**Challenge part 1**

Run a model for this and using the coefficients report the data. Remember they will be log odds.

<br>

<button type="button" class="btn btn-success">

Ensure you know what level the model is using as the baseline or reference level. It might be that it is useful to change this.  
</button>


```{r}
#| echo: false
#| eval: false
library(lme4)
m <- glmer(eat ~ group * sex + (1|lizard), data = lizards, family = binomial)
summary(m, corr = FALSE)
full.model <- glmer(eat ~ group * sex + (1|lizard), data = lizards, family = binomial)
reduced.model <- glmer(eat ~ 1 + (1|lizard), data = lizards, family = binomial)
anova(full.model, reduced.model)
```

<br>

**Challenge part 2**

Report the analysis in words as clearly as you can. Include a graph. Ask someone else if they can understand the results from what you have written.


<br>
<br>

# Nested design

Some experiments have what's called a nested design. For example the lizard experiment would be nested if the lizards had been housed in groups of 10 lizards to a tank rather than having a tank to themselves. There may be some differences among the tanks and therefore you would have to add `tank` into the model. Since not every individual is in every tank ie each one is only in one tank,  `lizard` is nested within `tank`. [nested lizard data](./data/lizard_nested.csv)

```{r}
lizards <- read.csv(file = "data/lizard_nested.csv")
head(lizards)
table(lizards$tank)
```


<br>

**Challenge part 1**

Find and run the code for this nested mixed model.

<button type="button" class="btn btn-success">

Tip - Identify the variables that are the **response**, the **fixed effect**, the **random effects** and what is nested in what.
</button>

<br>

**Challenge part 2**

Report the analysis. Include a graph. 

<br>

**Challenge part 3**

Decide if including the tank in the model changed the results?


```{r}
#| echo: false
#| eval: false

lizardsnest <- read.csv(file = "data/lizard_nested.csv")
model <- glmer(eat ~ 1 + (1|tank/lizard), data = lizardsnest, family = binomial)
summary(model)
```



```{r}
#| echo: false
#| eval: false
# from https://stackoverflow.com/questions/67123529/how-to-include-nested-effects-in-mixed-effect-model-with-lme4-package
mixed.lmer2 <- lmer(Y~ X + (1|mountain/site), data = data)
# for lizard example would be
m2 <- glmer(eat ~ 1 + (1|tank/lizard), data = lizards, family = binomial)
# classes and pupils
(1|class/pupil)  # or  
(1|class) + (1|class:pupil)
```



<br>

# Crossed random effects

Using the lizard examples again, a **crossed** design would be if the different lizards had experienced the different tanks (the researcher would have had to keep moving them around tanks). It would be a **partially crossed** design if each lizard had been measured in only some of the tanks and a **fully crossed** design if each lizard had been measured in all of the tanks.


<br>

**Challenge**

For each experiment below decide if it is a mixed model, nested, partially crossed or fully crossed design. 

1) Students are measured for statistics ability repeatedly in tests. The effects of gender and prior education are fixed effects. Students study different programmes.

2) Researchers test how light intensity (high vs low) and nutrient availability (high vs low) affect plant biomass across multiple forest sites. At each site, plots are divided into subplots that receive all combinations of light and nutrient treatments, creating a 2×2 factorial design.

3) Researchers want to test how soil type (sandy vs clay) influences plant biomass across different forest sites. They select 6 sites, and within each site, they identify 3 plots. In each plot, they measure biomass of 10 individual plants growing in the same soil type. 

```{r}
#| eval: false
#| echo: false
#1)mixed model with random as Student ID
#2)Fully crossed Sites and plots are treated as random effects, while light and nutrients are fixed factors, allowing the model to account for hierarchical sampling and interaction effects. This design ensures every level of one factor occurs with every level of the other across all plots.
#3) Nested Here, plants are nested within plots, and plots are nested within sites, while soil type is a fixed factor applied at the plot level. This design accounts for hierarchical structure and spatial dependence.
```




```{r}
#| eval: false
#| echo: false
# https://stats.stackexchange.com/questions/228800/crossed-vs-nested-random-effects-how-do-they-differ-and-how-are-they-specified
(1|class) + (1|pupil)
```

<br>

<br>
